Interpreting the gas-fraction as the entropy of an isothermal dusty gas

### 2016-07-28 model.tex
text and code update

\subsection{Locally isothermal equation of state for the gas} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \epsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time.  
%For analytic discussion we consider arbitrary functions of
%$c_s(r,z)$. 
In numerical calculations we will consider vertically  
isothermal disks with 
\begin{align}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature.

@@

\subsection{Disk structure with a prescribed dust distribution} 
We assume a Gaussian profile in the dust-to-gas ratio, 
\begin{align}\label{dust_gauss}
  \tepsilon(r,z) = \tepsilon_0
  \exp{\left(-\frac{z^2}{2\widetilde{H}^2}\right)}. 
\end{align}
For simplicity we assume the dust-to-gas ratio at the mid-plane
$\tepsilon_0$, as well as its characteristic scale-height
$\widetilde{H}$, are both constant.

% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \tepsilon_0(r) = \tepsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\tepsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.

Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  \rhog(r,z) = \rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\tepsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

@@

>>>>

\subsection{Locally isothermal equation of state for the gas} 
We adopt 
\begin{align}\label{eos}
  P = c_s^2(r,z)\rhog = c_s^{2}(1 - \epsilon)\rho,   
\end{align}
where $c_s$ is a prescribed sound-speed profile fixed in time.  
%For analytic discussion we consider arbitrary functions of
%$c_s(r,z)$. 
In numerical calculations we will consider vertically  
isothermal disks with 
\begin{align}\label{power_temp}
  c_s^2(r) \propto r^{q},
\end{align}
where $q$ is the power-law index for the disk temperature.

@@

\subsection{Disk structure with a prescribed dust distribution}\label{steady_state}  
We assume a Gaussian profile in the dust-to-gas ratio, 
\begin{align}\label{dust_gauss}
  \tepsilon(r,z) = \tepsilon_0(r)
  \exp{\left[-\frac{z^2}{2\widetilde{H}^2(r)}\right]}. 
\end{align}
%For simplicity we assume the dust-to-gas ratio at the mid-plane
%$\tepsilon_0$, as well as its characteristic scale-height
%$\widetilde{H}$, are both constant.
% We take the mid-plane dust-to-gas ratio
%to be a power-law in radius,  
%\begin{align}
%  \tepsilon_0(r) = \tepsilon_{00}\left(\frac{r}{r_0}\right)^{-d},  
%\end{align}
%where $\tepsilon_{00}$ is the dust-to-gas ratio at the fiducial radius
%$r_0$.  
Inserting Eq. \ref{dust_gauss} into vertical hydrostatic equilibrium,
Eq. \ref{steady_momz} and integrating with the approximate
gravitational potential (Eq. \ref{thin_disk_potential}) we obtain the
gas density as
\begin{align}
  \rhog(r,z) = \rho_\mathrm{g0}(r)\exp{\left\{ - \frac{z^2}{2\Hgas^2}
    -\tepsilon_0\frac{\Htilde^2}{\Hgas^2}\left[1 -
      \exp{\left(-\frac{z^2}{2\Htilde^2}\right)}\right] \right\}}, 
\end{align}
where
\begin{align}
  \Hgas = \frac{c_s}{\OmK}, \quad \OmK \equiv \sqrt{\frac{GM_*}{r^3}},   
\end{align}
is the gas scale-height in the dust-free limit and $\OmK$ is the
Keplerian frequency, respectively.

@@

Writing Eq. \ref{vshear} in terms of the gas density and dust-to-gas
ratio with a power-law temperature profil (Eq. \ref{power_temp}) e gives 
\begin{align*}
  &r\frac{\p \Omega^2}{\p z}  =
  \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{
  \frac{\p\tepsilon}{\p r}\frac{\p\ln{\rhog}}{\p z}
  -\frac{\p\tepsilon}{\p z}\frac{\p\ln{\rhog}}{\p r}\right.\notag\\
  &\phantom{ r\frac{\p \Omega^2}{\p z}  =
    \frac{c_s^2(r)}{\left(1+\tepsilon\right)^2}\left\{\right\} }
  \left. +\frac{q}{r}\left[\frac{\p\tepsilon}{\p z} + \left(1+\tepsilon\right)\frac{\p\ln{\rhog}}{\p z}\right]
  \right\} 
\end{align*}
The first two terms correspond to vertical shear caused by spatial
variations in the dust-to-gas ratio; while the second term
proportional to $q$ corresponds to vertical shear due to the 
radial temperature gradient.

We can compare these sources by
evaluating them, to order-of-magnitude, using the equilibrium
solutions in \S\ref{steady_state}. We assume the disk is radially
smooth so that $\p_r\sim 1/r$, and the dust-to-gas ratio
$\tepsilon\ll1$. This gives

\begin{align}\label{vshear_split}
  \mathcal{V} =  \frac{\left|r\p_z\Omega\right|_{\text{
        dust/gas gradient}}}{\left|r\p_z\Omega\right|_{\text{
        temp. gradient}}} \sim
  \tepsilon \frac{\mathrm{max}\left(\delta^2,
    1\right)}{\left|q\right|\left(\tepsilon + \delta^2\right)},
\end{align}
where $\delta\equiv \Htilde/\Hgas$. 
Since $|q|=O(1)$ in PPDs, Eq. \ref{vshear_split} indicates that
vertical shear due to variations in the dust-to-gas ratio can be at
most comparable to that due to the radial temperature gradient; and
this only occurs for thin dust layers with $\delta^2\ll \tepsilon$ (so
that $\mathcal{V}\sim 1/|q|$). Otherwise, vertical shear is
predominantly due to $\p_rc_s^2$.
